---
title: An R Markdown document converted from "hw2_Applying_Fun_R.ipynb"
output: html_document
---


# Applying functions in R

```{r}
options(repos = c(CRAN = 'https://cloud.r-project.org'))
```


```{r} 
install.packages('rio')#adding the necessary package to read the data.
```

```

# reading data

```{r}
# opening a file from a url (file in Excel), name it 'fragility23'
linkGit="https://github.com/Fundamentals-Josh/hw2/raw/refs/heads/main/FSI-2023-DOWNLOAD.xlsx"  #modified code structure to point to the file I have added in my own repository. Access and read the data from the file. 

# library(rio) # package needed
fragility23=rio::import(file = linkGit) #object that will hold the result
```

```{r}
str(fragility23) #displays internal data structure of the data frame being examined.
```

## Apply square root function?


# whole DF?
sqrt(fragility23) #using this command attempts to obtain the square root of the entire data function.  As the data frame includes non-numeric variables, use of this command triggers an error code.
```

```{r}
# several columns (only numerical)

sqrt(fragility23[,4:5]) #this function limits the square root function to selected columns.  In this case, the function successfully executes, because the selected columns contains numerical data.
```

```{r}
# one column
sqrt(fragility23$Total)  #this function finds the square root of the 'Total' column of the data frame for each row.
```

```{r}
# one value
sqrt(fragility23$Total[1]) #this function finds the square root of row 1 of the total column in the data frame. 
```

## Applying **sum()**:

```{r}
# sum everything
sum(fragility23[,4:5]) #this function finds the total sum of columns 4 and 5 of the data frame. 
```

```{r}
# sum by column
print(apply(fragility23[,4:5],2,sum)) #this function sums columns 4 and 5, but as separate figures.
```

```{r}
#notice
typeof(apply(fragility23[,4:5],2,sum)) #this function provides you with the data type for the previous command.  As the totals include decimals, the data type is a double. 
```

If you do not see **list**, then it is a vector. ⏫

```{r}
# sum by row
print(apply(fragility23[,4:5],1,sum)) #this function applies the sum by row.
```

### Apply by iterating:

```{r}
print(lapply(fragility23[,4:5],sum)) #this function will sum separately by column, but will treat the data as a list.  
```

Notice output of **lapply**:

```{r}
typeof(lapply(fragility23[,4:5],sum)) #shows the datatype of the previous function.  This command confirms that the lapply function results in the data being treated as a list, rather than a vector.
```

```{r}
class(lapply(fragility23[,4:5],sum))  #this function confirms that the data class of the function is a list.
```

Notice output of **sapply**:

```{r}
print(sapply(fragility23[,4:5],sum)) #this function provides separate sums for columns 4 & 5, but treats the data class as numeric.
```

```{r}
class(sapply(fragility23[,4:5],sum)) #this function confirms the stated function treats the class for the data as numeric
```

Similarly:

```{r}
print(lapply(fragility23[,4:5],sqrt))  #this function figures the square root for the figures in columns 4 and 5 but treats the data class as a list.
```

```{r}
class(lapply(fragility23[,4:5],sqrt)) #this function confirms the previous function is treating the data from columns 4 & 5 as a list. 
```

```{r}
print(sapply(fragility23[,4:5],sqrt)) #this function figures the square root for figures 4 & 5.
```

```{r}
class(sapply(fragility23[,4:5],sqrt)) #this function confirms that the data is being treated as a matrix array (vector).  
```

Now our own function:

```{r}
theOnesOK = function(DF_country_and_variable) { #This defines a function named theOnesOK that accepts one argument: DF_country_and_variable, which is expected to be a data frame.
  variable_values <- DF_country_and_variable[,2]#It extracts all values from the second column of the input data frame and stores them in a vector named variable_values. 
  avg_value <- mean(variable_values, na.rm = TRUE) #It calculates the mean (average) of the variable_values vector. na.rm = TRUE ensures that any missing values (NA) are ignored during the mean calculation.
  is_above <- ifelse(variable_values > avg_value, "Above Average", "Below/At Average") #This line creates a new vector called is_above. For each value in variable_values, it checks if that value is greater than the calculated avg_value. If it is, the corresponding element in is_above becomes "Above Average"; otherwise, it becomes "Below/At Average".
  DF_country_and_variable$Status <- is_above #It adds a new column named Status to the original input data frame DF_country_and_variable and populates it with the values from the is_above vector.
  return(DF_country_and_variable) #The function returns the modified data frame, which now includes the new Status column.
}
``` 

```{r}
theOnesOK(fragility23[,c('Country','S1: Demographic Pressures')]) #This takes the previously constructed command and applies it to the defined columns of the data frame. 
```



```{r}
mystery=function(DF,positionsToUse,CountryColumn='Country'){# This defines a function named mystery that has three components: DF The input data frame. positionsToUse: A vector of column indices from the data frame where we want to calculate the average. CountryColumn: The name of the column in DF that contains country information (defaults to 'Country').

  newDF=DF[,c(CountryColumn),drop = FALSE] #It creates a new data frame called newDF containing only the CountryColumn from the original DF
  average='average' #defines the label 'average' to be used as the name for the new column that will store the calculated averages.
  newDF[,average]=apply(DF[,positionsToUse],1,mean,na.rm = TRUE)#creates a new data frame called newDF containing only the CountryColumn from the original DF
  return(newDF[,c(CountryColumn,average)]) #the function returns the newDF data frame, but it explicitly selects only the CountryColumn and the newly created 'average' column to be part of the output.
}
```

```{r}
mystery(fragility23,4:6) #applies the previous command from columns 4 to 6
```

```{r}
theOnesOK2 = function(DF, DFvariable, CountryColumn='Country') { #This defines a function named theOnesOK2 and has three components, DF: The input data frame, DFvariable: A string representing the name of the column in the data frame that contains the numerical values,CountryColumn, The name of the column in DF that contains country information.

  variable_values <- DF[,DFvariable] #It extracts all values from the column specified by DFvariable from the input DF and stores them in a vector named variable_values.
  avg_value <- mean(variable_values, na.rm = TRUE) #It calculates the mean (average) of the variable_values vector. na.rm = TRUE ensures that any missing values (NA) are ignored during the mean calculation.
  is_above <- ifelse(variable_values > avg_value, "Above Average", "Below/At Average") #This line creates a new vector called is_above. For each value in variable_values, it checks if that value is greater than the calculated avg_value. If it is, the corresponding is in is_above becomes "Above Average"; otherwise, it becomes "Below/At Average".
  newname = paste('Status_on', DFvariable) #This command creates a name for the new status column, depending on the column selected. For example, if DFvariable is 'S1: Demographic Pressures', newname will become 'Status_on S1: Demographic Pressures'.
  DF[,newname] <- is_above #Adds a new column named newname (per the above command) to the original input data frameand populates it with the values from the is_above vector.
  return(DF[,c(CountryColumn, newname)]) # The function returns a new data frame that contains only the CountryColumn and the newly created status column (newname) from the modified data fr.
}
```

```{r}
theOnesOK2(fragility23, 'S1: Demographic Pressures')#executes above command. 
```

